---
title: "Best Practise in R"
author: "Gareth Burns"
date: "09/12/2021"
output: 
  ioslides_presentation:
    widescreen: true
    logo: ../img/exploristics.jpg
    css: styles.css
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(palmerpenguins)
library(ggplot2)

```

## Best Practise in R

R is an open source ecosystem with vast array of different packages from
many different authors, evolving of over many years and across multiple
different disciples. There are no agreed conventions, even within base
R. See `?fisher.test()` as an example.

This makes it difficult for a R user to know what convention to adopt.

**Aim:** To *stimulate* discussion on best practise, *signpost* to
resources & *suggest* some approaches to take.

## Key messages

The key take home messages from the *Jumping Rivers* course were:

-   Write code for **humans** to understand, not computers
-   Best practise is subjective and specific to individual use-cases
-   Adopt conventions, document them and be **consistent** within a Team
-   Write a coding style guide
-   Adopting best practice may not save you time short term but will
    save time in long term
-   There's multiple things to consider when writing code and no one
    right way (trade-off)

## Naming Conventions

Code will run without a standard naming convention but it will make it
difficult for others to read and interpret the code. It can lead to
bugs, difficulty to communicate what the code does and inefficiencies.

In KerusCloud this has caused bugs when communicating between Database
API R-engine with similar but slightly different naming conventions.

Some things we name in R: - Files - Functions - Arguments - Variables -
Classes

## Naming Conventions

```{r naming_conventions, eval=TRUE}
namingConventions <- read.csv(file = "../data/NamingConvention.csv")

knitr::kable(namingConventions, "html")
```

## Common Style Guidelines

-   Put space after a *comma* and not before
-   Put space before and after an *equals* sign
-   Use assignment operator (`<-`) for created functions and variables
-   Maximum 80 characters per line
-   Indent curly brackets (RStudio: *Crtl + Shift + A*)
-   Don't leave in commented out code - hard to know why it's commented
    out
-   Read Tidyverse style guide

## Useful tools

-   Janitor package (cleaning up imported names)
-   Code Formatters (Format the code for you)
-   Code Linters (Tell you if you're following style guidelines)
-   R Studio Add-Ins

## Modularise Code

Two key elements

1.  Splitting into smaller components (e.g. functions)
2.  Collecting components together (e.g. project or package)

**Don't repeat yourself** (DRY). If you're copying and pasting code it's
a sign you should be writing a function. If the code is in several
locations then if you need to update it, you will need to update it in
several locations.

## Modularise Code

Code can evolve in *spaghetti code* overtime due to *ad-hoc* changes and
poor planning & standards. This can be difficult to read, hard to
identify issues and does not scale well. Modularising your code solves
these issues and has multiple advantages:

-   Easy to re-use components elsewhere (**resource**)
-   Navigate code-base easier
-   Different team members can work on different independent components
-   Scales better
-   Facilitates adopting of pre-existing solutions for components
-   Testing and validation can occur at a lower granularity,
    pin-pointing problems

## Other Topics

There isn't enough time to go over other topics but there's some great
material in Jumping Rivers notes and each could be a Thursday topic.

-   Functions (**page 12**)
-   RStudio Projects (**4.1 page 26**)
-   Dependency Management (**4.5 page 28**)
-   Version Control (**4.6 page 29**)
-   R packages (**4.7 page 29**)

## Lessons Learnt 1

The precedence of a function will depend on the order you load packages.
You can call a function from a specific package using the syntax. \>
package_name::function_name

For example, both the *dplyr* and *MASS* package have the function
`select` and R will use the `select` function from the last loaded
package.

```{r PackageOrder, echo=TRUE, eval=FALSE}
library(dplyr)
library(MASS)

# This code will return an error
penguins %>%
 select(species, island, bill_length_mm)

# This code will run
penguins %>%
 dplyr::select(species, island, bill_length_mm)
```

## Lessons Learnt 2

Don't *hard-code* variables, create a variable at start of script.

```{r palmer_penguins_code, echo=TRUE, eval=FALSE}
gentooData <- penguins[penguins$species == "Gentoo", ]

ggplot(data = gentooData,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3) +
      labs(
        title = "Relationship between flipper length and body mass for Gentoo Penguins",
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)",
      ) +
      theme_minimal()
```

## Lessons Learnt 2

```{r palmer_penguins_plot, echo=FALSE, eval=TRUE, warning=FALSE}
gentooData <- penguins[penguins$species == "Gentoo", ]

ggplot(data = gentooData,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3 ) +
      labs(
        title = "Relationship between flipper length and body mass for Gentoo Penguins",
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)",
      ) +
      theme_minimal(
      )
```

## Lessons Learnt 2

Copy and pasting code is a common practice but but can lead to errors
that aren't always picked up on.

```{r palmer_penguins_code_error, echo=TRUE, eval=FALSE}
chinstrapData <- penguins[penguins$species == "Chinstrap", ]

ggplot(data = chinstrapData,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3) +
      labs(
        title = "Relationship between flipper length and body mass for Gentoo Penguins",
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)"
      ) +
      theme_minimal()
```

## Lessons Learnt 2

```{r palmer_penguins_plot_error, echo=FALSE, eval=TRUE, warning=FALSE}
chinstrapData <- penguins[penguins$species == "Chinstrap", ]

ggplot(data = chinstrapData,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3 ) +
      labs(
        title = "Relationship between flipper length and body mass for Gentoo Penguins",
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)"  ) +
      theme_minimal()
```

## Lessons Learnt 2

Create a variable at the start of the script mitigates this risk and
makes it easier to change the scope of the code.

```{r palmer_penguins_plot_suggested, echo=FALSE, eval=TRUE, warning=FALSE}
species <- "Adelie"
dataset <- penguins[penguins$species == species, ]

ggplot(data = dataset,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3 ) +
      labs(
        title = paste("Relationship between flipper length and body mass for", species ,"Penguins"),
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)",
      ) +
      theme_minimal()
```

## Lessons Learnt 3

Create a script to carry out as much tasks as possible or document where
externalities occur. Ideally if you're obtaining external data, write a
script for sourcing the data and save it in the same project file
structure. If the data was obtained via USB, e-mail or one off transfer
- document this. As the next person coming along will have no way of
knowing where these data were obtained from!

## Lessons Learnt 4

When creating an index for referencing (such as a `for` loop or
`lapply`), use **`seq_len`**. Other sequence index creating methods can
create a sequence backwards and make it difficult to debug loops.

```{r seq_generation, echo=TRUE}
negativeNumber <- -4
zero <- 0

1:negativeNumber
1:zero

seq(negativeNumber)
seq(zero)
```

```{r seq_len, echo=TRUE, eval=FALSE}
# These will return an error before your loop is run
seq_len(negativeNumber)
seq_len(zero)
```

## Lessons Learnt 5

Use **`TRUE`** & **`FALSE`** not **`T`** & **`F`**. **`TRUE`** &
**`FALSE`** are protected terms in R. **`T`** & **`F`** are not.

```{r TrueFalse, echo = TRUE}
T <- FALSE
# NA's propagate
mean(penguins$bill_length_mm)

# Change the na.rm argument to TRUE by using shorthand T
mean(penguins$bill_length_mm, na.rm = T)
isFALSE(T)
mean(penguins$bill_length_mm, na.rm = TRUE)
```

## Lessons Learnt 5

```{r TrueFalse2, echo = TRUE, eval=FALSE}
# TRUE is a protected term in R
TRUE <- FALSE
```

```{r TrueFalse3, echo = TRUE, eval=TRUE}
mean(penguins$bill_length_mm, na.rm = TRUE)
```

## Lessons Learnt 6

Running a script will automatically overwrite your output files! If you
copy and paste a script and start to modify it this can lead to
unintended consequences! A simple wrapper function can check if a file
exists and if so append a date stamp suffix to prevent overwriting.

For example of a wrapper function with Roxygen comments see

## Lessons Learnt 7

Use **`isTRUE`** or **`isFALSE`** for conditionals (i.e. if statements).
If you supply more than one logical argument to an if statement it will
take the first argument and display a warning.
