---
title: "Best Practise in R"
author: "Gareth Burns"
date: "02/12/2021"
output: 
  ioslides_presentation:
    widescreen: true
    logo: ../img/exploristics.jpg
    css: styles.css
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(palmerpenguins)
library(ggplot2)

```

## Best Practise in R

R is an open source ecosystem with vast array of different packages from many different authors, 
evolving of over many years and across multiple different disciples. There are no agreed conventions, 
even within base R.

This makes it difficult for a R user to know what convention to adopt.

*Aim:* To *stimulate* discussion on best practise, *signpost* to resources & *suggest*
some approaches to take. 

## Key messages

The key take home messages from the *Jumping Rivers* course were:

- Write code for **humans** to understand, not computers
- Best practise is subjective and specific to individual use-cases
- Adopt conventions, document them and be *consistent* within a Team
- Adopting best practice may not save you time short term but will save time in long term
- There's multiple things to consider when writing code and no one right way (trade-off)


## Naming Conventions 

Code will run without a standard naming convention but it will make it 
difficult for others to read and interpret the code. It can lead to bugs,
difficulty to communicate what the code does and inefficiencies. In Kerus this
has caused bugs between UI-Database-API-R engine with similar but slightly different 
naming.
Some things we name in R:
- Files
- Functions
- Arguments
- Variables
- Classes

## Naming Conventions Examples



## Common Style Guidelines

- Put space after a *comma* and not before
- Put space before and after an *equals* sign 
- Use assignment operator (<-) for created functions and variables
- Maximum 80 characters per line
- Indent curly brackets (RStudio: *Crtl + Shift + A*)


## Modularise Code

Two key elements  

1.  Splitting into smaller components (e.g. functions)
2.  Collecting components together (e.g. project or package)

**Don't repeat yourself** (DRY). If you're copying and pasting code it's a sign 
you should be writing a function. If the code is in several locations then if
you need to update it, you will need to update it in several locations.


## Modularise Code

Code can evole in *spaghetti code* overtime due to ad-hoc changes and poor planning 
& standards. This can be difficult to read, hard to identify issues and does not scale.
Modularising your code solves these issues and has multiple advantages


## Lessons Learnt 1 

The precedence of a function will depend on the order you load packages. You can
call a function from a specific package using the syntax 
> package_name::function_name

```{r PackageOrder, echo=TRUE, eval=FALSE}
library(dplyr)
library()

```

## Lessons Learnt 2

Don't *hard-code* variables, create a variable at start of script.

```{r palmer_penguins_code, echo=TRUE, eval=FALSE}
library(palmerpenguins)
library(ggplot2)

gentooData <- penguins[penguins$species == "Gentoo", ]

ggplot(data = gentooData,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3 ) +
      labs(
        title = "Relationship between flipper length and body mass for Gentoo Penguins",
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)",
      ) +
      theme_minimal(
      )
```

## Lessons Learnt 2

```{r palmer_penguins_plot, echo=FALSE, eval=TRUE}
library(palmerpenguins)
library(ggplot2)

gentooData <- penguins[penguins$species == "Gentoo", ]

ggplot(data = gentooData,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3 ) +
      labs(
        title = "Relationship between flipper length and body mass for Gentoo Penguins",
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)",
      ) +
      theme_minimal(
      )
```

## Lessons Learnt 2

Copy and pasting code is a common practice but but can lead to errors that aren't 
always picked up on. 

```{r palmer_penguins_code_error, echo=TRUE, eval=FALSE}
library(palmerpenguins)
library(ggplot2)

chinstrapData <- penguins[penguins$species == "Chinstrap", ]

ggplot(data = chinstrapData,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3 ) +
      labs(
        title = "Relationship between flipper length and body mass for Gentoo Penguins",
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)"
      ) +
      theme_minimal()
```

## Lessons Learnt 2

```{r palmer_penguins_plot_error, echo=FALSE, eval=FALSE}


chinstrapData <- penguins[penguins$species == "Chinstrap", ]

ggplot(data = chinstrapData,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3 ) +
      labs(
        title = "Relationship between flipper length and body mass for Gentoo Penguins",
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)"  ) +
      theme_minimal()
```

## Lessions Learnt 2

Create a variable at the start of the script mitigates this risk and makes it 
easier to change the scope of the code. 

```{r palmer_penguins_plot_suggested, echo=FALSE, eval=TRUE}
library(palmerpenguins)
library(ggplot2)

species <- "Adelie"
dataset <- penguins[penguins$species == species, ]

ggplot(data = dataset,
       aes(x = flipper_length_mm,
           y = body_mass_g)) +
  geom_point(aes(color = "Red"),
    size = 3 ) +
      labs(
        title = paste("Relationship between flipper length and body mass for", species ,"Penguins"),
        caption = "Data Source: Palmers Penguins",
        x = "Flipper length (mm)",
        y = "Body mass (g)",
      ) +
      theme_minimal()
```


## Lessons Learnt 3

Create a script to carry out as much tasks as possible or document where externalities 
occur. Ideally if you're obtaining external data, write a script for sourcing the data
and save it in the same project file structure. 
If the data was obtained via USB, e-mail or one off transfer - document this. As 
the next person coming along will have no way of knowing where these data were 
obtained from!

## Lessons Learnt 4

When creating an index for referencing (such as a loop or lapply), use **seq_len**.
Other sequence index creating methods can create a sequence backwards and 
make it difficult to debug loops.

```{r seq_generation, echo = TRUE}
negativeNumber <- -4
zero <- 0

1:negativeNumber
1:zero

seq(negativeNumber)
seq(zero)
```

```{r seq_len, echo = TRUE, eval=FALSE}
# These will return an error before your loop is run
seq_len(negativeNumber)
seq_len(zero)
```

## Lessons Learnt 5 

Use **TRUE** & **FALSE** not **T** & **F**.  **TRUE** & **FALSE** are protected
terms in R. **T** & **F** are not. 

```{r TrueFalse, echo = TRUE}
T <- FALSE

# NA's propagate
mean(penguins$bill_length_mm)

# Change the na.rm argument to TRUE by using shorthand T
mean(penguins$bill_length_mm, na.rm = T)

isFALSE(T)

mean(penguins$bill_length_mm, na.rm = TRUE)
```

```{r TrueFalse2, echo = TRUE, eval = FALSE}
# TRUE is a protected term in R
TRUE <- FALSE
```

## Lessons Learnt 6

Running a script will automatically overwrite your output files! If you copy and
paste a script and start to modify it this can lead to unintended consequences!
A simple wrapper function can check if a file exists and if so append a date stamp 
suffix to prevent overwriting.

## Lessons Learnt 7

Use **isTRUE** or **isFALSE** for conditionals (e.g. if statements). If you supply 
more than one logical argument to an if statement it will take the first argument. 

## Resources

- Jumping Rivers Course Notes
- Tidyverse style guide <https://style.tidyverse.org/>
- Existing Packages
- Each other

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

